An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2022-03-24 submissions/submission-ju-hofmann/ju-hofmann.tex
reorganize the folder

In this paper we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually describe the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \pkg{cmpsR} on CRAN.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in \autoref{fig:bullet}. During the firing process,
grooves and lands are engraved on a bullet. Firearms examiners use
striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to
\href{http://sourceforge.net/p/open-gps/mwiki/X3p/}{the ISO5436-2
standard} adopted by the OpenFMC (Open Forensic Metrology Consortium), a
group of firearm forensics researchers who contributes to the
establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification
with the scan resolution being 1.5625 \textmu m \(\times\) 1.5625
\textmu m per pixel. Hamby set 44 was scanned at the Roy J Carver
High-Resolution microscopy lab at Iowa State. These scans were acquired
with a Sensofar Confocal Light Microscope at 20x magnification for a
nominal resolution of 0.645 \textmu m \(\times\) 0.645 \textmu m per
pixel. Both Hamby set 252 and Hamby set 44 are publicly available from
the NIST Ballistics Database Project \citep{nistdb}.

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates
which cross-section was identified by the algorithm to represent the
LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional
view. Groove engraved areas are removed from the analysis as indicated
by the vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found in the ``Supplementary materials''. Note
that the process of extracting signatures might be different from the
one used in \citet{cmps} because no code or parameter settings are made
available publicly.

@@

Most algorithms for comparing striation marks are based on the digitized
signatures and produce a similarity score
\citep[\citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{song2005}. The congruent matching profile segments
(CMPS) algorithm, developed by \citet{cmps} for ``objective comparison
of striated tool marks'', is one such algorithm. The algorithm's main
idea is to take a set of consecutive and non-overlapping basis segments
from the comparison and for each segment find the ``best'' registration
position on the reference (the other bullet signature) with respect to
their cross-correlation values. From a comparison of these registration
positions, a \textbf{congruent registration position} is identified, and
the number of basis segments taking the congruent registration position
is the CMPS score. Note that researchers in \citet{cmps} refer to the
origin of basis segments as the reference, but in this paper we refer to
it as the comparison signature. High CMPS scores are achieved between
more similar signatures and are therefore indicative of a same-source
pair. Low scores between pairs of signatures are attributed to different
source pairs. However, a specific threshold of the CMPS score to
distinguish between same-source and different-source comparisons is not
provided in \citet{cmps}, instead the threshold depends on the
underlying structure of the data and the choice of parameters. In a
legal setting this variability is problematic because it allows for
situations in which experts could choose parameters based on whether
they are witnesses for the defense or the prosecution. Further research
is needed to understand how to determine optimal threshold settings. The
CMPS algorithm can assist firearm examiners with drawing a conclusion
about the source of a comparison pair. Thus, in this paper we present an
open-source implementation of the CMPS algorithm in the R package
\pkg{cmpsR} available from both CRAN and Github. This publicly available
implementation calculates the CMPS score of a comparison using the
following code:

@@

As presented in the work of \citet{cmps}, researchers applied the CMPS
method to scans of one of the Hamby sets. While it is not explicitly
stated in the paper, we presume this to be Hamby 252, as only those
scans were publicly available at the time. In order to show that our
implementation of the CMPS algorithm is able to reproduce the results in
\citet{cmps} and be used for other data sets, we applied our
implementation to both Hamby set 252 and Hamby set 44. For the purpose
of reproducibility, we present how we obtained bullet signatures from
the Hamby set data and include code in the appendix: for both Hamby 252
and Hamby 44, we started with scans in the form of x3p files in the
database. Following the framework proposed by \citet{aoas}, we used the
same set of parameters, removed damaged bullet scans, obtained bullet
signatures for each bullet land engraving, and removed outliers in
bullet signatures. Note that researchers of \citet{cmps} applied the
CMPS algorithm to bullet signatures as well but used a framework
different from ours. However, since their work is not open-source, we
were not able to follow their framework and were only able to reproduce
the results for Hamby set 252 qualitatively.

@@

For details about the implementation, examples and results, please refer
to the ``Supplementary materials''.

@@

\hypertarget{supplementary-materials}{%
\subsection{Supplementary materials}\label{supplementary-materials}}

The zip file ``supplementary-files.zip'' contains R scripts for
reproducing all aspects of this paper.

\textbf{Data not included in ``supplementary-files.zip''}

The processed data of Hamby set 252 and Hamby set 44 are stored in two
\texttt{.rds} files:

\begin{itemize}
\tightlist
\item
  \href{https://github.com/willju-wangqian/CMPSpaper/blob/main/reproducible/bullet_signatures_etc/BulletSignatures252.rds}{\texttt{BulletSignatures252.rds}}
\item
  \href{https://github.com/willju-wangqian/CMPSpaper/blob/main/reproducible/bullet_signatures_etc/BulletSignatures44.rds}{\texttt{BulletSignatures44.rds}}
\end{itemize}

And the ground truth of Hamby set 252 is provided in
\href{https://github.com/willju-wangqian/CMPSpaper/blob/main/reproducible/bullet_signatures_etc/StudyInfo.xlsx}{\texttt{StudyInfo.xlsx}}

Due to the size of the file, these processed data are not included in
the ``supplementary-files.zip'', but the links are provided for
download.

In order to reproduce the results in the paper, please download
``BulletSignatures252.rds'', ``BulletSignatures44.rds'', and
``StudyInfo.xlsx'' and save them in a folder named
``bullet\_signatures\_etc''. And place the folder
``bullet\_signatures\_etc'' into the folder of all the R scripts of
``supplementary-files.zip''.

Please check out the folder structure of
\href{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}{the
\texttt{reproducible} folder} for reference.

\textbf{File description}

\begin{itemize}
\tightlist
\item
  \texttt{hamby*\_result\_generator.R}: these R scripts take the
  processed data of Hamby set 252 and Hamby set 44, generate preliminary
  results used in the paper, and save the results in the \texttt{.csv}
  format in the folder \texttt{data-csv}.
\item
  \texttt{rds\_generator.R}: this R script takes the generated
  \texttt{.csv} files and produce \texttt{CMPSpaper\_results.rds}.
  \texttt{CMPSpaper\_results.rds} is identical to
  \texttt{data/CMPSpaper\_results.rds} and is used generate figures and
  other results presented in the paper.
\item
  \texttt{func\_collection.R}: this R script contains some helper
  functions used by other R scripts.
\end{itemize}

\textbf{Data included in ``supplementary-files.zip''}

\begin{itemize}
\tightlist
\item
  \texttt{data-csv}: this folder contains \texttt{.csv} files we
  generated using R scripts \texttt{hamby*\_result\_generator.R}
\item
  \texttt{CMPSpaper\_results.rds}: this is the \texttt{.rds} file we
  generated using R script \texttt{rds\_generator.R}
\end{itemize}

These data can be used as reference or example results of the
reproducible codes

\begin{itemize}
\tightlist
\item
  \texttt{csafe\_rf2.rds}: this \texttt{.rds} file contains the random
  forest model used in this paper
\end{itemize}

@@

\address{%
Wangqian W. Ju\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/willju-wangqian}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-9977-377X}{0000-0002-9977-377X}}\\%
\href{mailto:wju@iastate.edu}{\nolinkurl{wju@iastate.edu}}%
}

\address{%
Heike Hofmann\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/heike}\\%
\textit{ORCiD: \href{https://orcid.org/0000-0002-9079-593X}{0000-0002-9079-593X}}\\%
\href{mailto:hofmann@iastate.edu}{\nolinkurl{hofmann@iastate.edu}}%
}

>>>>

In this paper we present an open-source implementation of the algorithm
of the Congruent Matching Profile Segments (CMPS) method. \citet{cmps}
developed the CMPS method for ``objective comparison of striated tool
marks'' and demonstrated its use in some examples of comparing bullet
signature correlations. Although \citet{cmps} conceptually describe the
CMPS algorithm in their paper, the authors did not release an
implementation of their method. Thus, our effort here is to introduce
the CMPS method to the \texttt{R} \citep{R} community and provide an
open-source, publicly available implementation of the algorithm to use,
review, and improve. Our implementation is made available as part of the
\texttt{R} package \pkg{cmpsR} on GitHub at
\url{https://github.com/willju-wangqian/cmpsR}.

@@

The datasets we worked with come from the James Hamby Consecutively
Rifled Ruger Barrel Study \citep{brundage, hamby, Hamby:2019}, in
particular, Hamby set 252 and Hamby set 44. For each Hamby set, a total
of 35 bullets is fired from (the same) ten consecutively manufactured
Ruger P-85 pistol barrels. Two bullets are fired from each barrel,
making up a set of 20 reference bullets. An additional 15 bullets are
fired from these ten barrels in a fashion unknown to the study
participant. The aim of the Hamby Study was to have firearms examiners
identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves
and lands as shown in \autoref{fig:bullet}. During the firing process,
grooves and lands are engraved on a bullet. Firearms examiners use
striation marks on land engraved areas (LEAs) for their visual
comparison. For algorithmic purposes, 3D topographical images of land
engraved areas were obtained and stored in x3p format (XML 3-D Surface
Profile). The x3p format provides a standard way of exchanging 2D and 3D
profile data. It conforms to the ISO5436-2 standard
{[}\url{http://sourceforge.net/p/open-gps/mwiki/X3p/}{]} adopted by the
OpenFMC (Open Forensic Metrology Consortium), a group of firearm
forensics researchers who contributes to the establishment of best
practices of using metrology in forensic science. Hamby set 252 was
scanned using a NanoFocus lens at 20x magnification with the scan
resolution being 1.5625 \textmu m \(\times\) 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy
lab at Iowa State. These scans were acquired with a Sensofar Confocal
Light Microscope at 20x magnification for a nominal resolution of 0.645
\textmu m \(\times\) 0.645 \textmu m per pixel. Both Hamby set 252 and
Hamby set 44 are publicly available from the NIST Ballistics Database
Project \citep{nistdb}.

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates
which cross-section was identified by the algorithm to represent the
LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional
view. Groove engraved areas are removed from the analysis as indicated
by the vertical blue lines in \autoref{fig:process}(c) and
\autoref{fig:process}(d)). A non-parametric LOESS smooth \citep{loess}
is fitted to capture the bullet curvature (\autoref{fig:process}(e))
and, finally, the \textbf{bullet signature} (\autoref{fig:process}(f))
is obtained as residuals of the cross-section and the fitted smooth.
Note, that in \citet{cmps} bullet signatures are referred to as bullet
profiles. However, to avoid confusion, we distinguish the notion of
bullet signatures from bullet profiles. Bullet profiles are shown in
panels (b), (c), and (d) of \autoref{fig:process}, while
\autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a
correct down-stream analysis of the signatures. To allow for a
human-in-the middle inspection and intervention we have provided an
interactive web-application, implemented as \texttt{R} Shiny App
\citep{shiny}, \pkg{bulletinspectR} to identify and correct those
errors. An example of the extraction process with corresponding code and
parameter settings can be found on Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}.
Note that the process of extracting signatures might be different from
the one used in \citet{cmps} because no code or parameter settings are
made available publicly.

@@

Most algorithms for comparing striation marks are based on the digitized
signatures and produce a similarity score
\citep[\citet{ChumbleyL_Scott2010VoTM}, \citet{aoas},
\citet{pmid30444940}]{song2005}. The congruent matching profile segments
(CMPS) algorithm, developed by \citet{cmps} for ``objective comparison
of striated tool marks'', is one such algorithm. The algorithm's main
idea is to take a set of consecutive and non-overlapping basis segments
from the comparison and for each segment find the ``best'' registration
position on the reference (the other bullet signature) with respect to
their cross-correlation values. From a comparison of these registration
positions, a \textbf{congruent registration position} is identified, and
the number of basis segments taking the congruent registration position
is the CMPS score. Note that researchers in \citet{cmps} refer to the
origin of basis segments as the reference, but in this paper we refer to
it as the comparison signature. High CMPS scores are achieved between
more similar signatures and are therefore indicative of a same-source
pair. Low scores between pairs of signatures are attributed to different
source pairs. However, a specific threshold of the CMPS score to
distinguish between same-source and different-source comparisons is not
provided in \citet{cmps}, instead the threshold depends on the
underlying structure of the data and the choice of parameters. In a
legal setting this variability is problematic because it allows for
situations in which experts could choose parameters based on whether
they are witnesses for the defense or the prosecution. Further research
is needed to understand how to determine optimal threshold settings. The
CMPS algorithm can assist firearm examiners with drawing a conclusion
about the source of a comparison pair. Thus, in this paper we present an
open-source implementation of the CMPS algorithm in the R package
\pkg{cmpsR} available from both CRAN and Github
\url{https://github.com/willju-wangqian/cmpsR}. This publicly available
implementation calculates the CMPS score of a comparison using the
following code:

@@

As presented in the work of \citet{cmps}, researchers applied the CMPS
method to scans of one of the Hamby sets. While it is not explicitly
stated in the paper, we presume this to be Hamby 252, as only those
scans were publicly available at the time.
(\hh{We have asked the authors of}
\citet{cmps}\hh{, and we are still waiting for the replies.}) In order
to show that our implementation of the CMPS algorithm is able to
reproduce the results in \citet{cmps} and be used for other data sets,
we applied our implementation to both Hamby set 252 and Hamby set 44.
For the purpose of reproducibility, we present how we obtained bullet
signatures from the Hamby set data and include code in the appendix: for
both Hamby 252 and Hamby 44, we started with scans in the form of x3p
files in the database. Following the framework proposed by \citet{aoas},
we used the same set of parameters, removed damaged bullet scans,
obtained bullet signatures for each bullet land engraving, and removed
outliers in bullet signatures. Note that researchers of \citet{cmps}
applied the CMPS algorithm to bullet signatures as well but used a
framework different from ours. However, since their work is not
open-source, we were not able to follow their framework and were only
able to reproduce the results for Hamby set 252 qualitatively.

@@

The code generating all results of this manuscript can be found on
Github at
\url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}

@@

@@

\address{%
Wangqian W. Ju\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/willju-wangqian}%
\\\textit{ORCiD: \href{https://orcid.org/0000-0002-9977-377X}{0000-0002-9977-377X}}%
\\\href{mailto:wju@iastate.edu}{\nolinkurl{wju@iastate.edu}}
}

\address{%
Heike Hofmann\\
Department of Statistics\\%
Center for Statistics and Applications in Forensic Evidence\\ Iowa State
University\\ 2438 Osborn Dr\\ Ames, IA 50011\\
%
\url{https://github.com/heike}%
\\\textit{ORCiD: \href{https://orcid.org/0000-0002-9079-593X}{0000-0002-9079-593X}}%
\\\href{mailto:hofmann@iastate.edu}{\nolinkurl{hofmann@iastate.edu}}
}
